\documentclass[runningheads]{llncs}

\usepackage{xspace}
\usepackage{proof}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{cmll}
\usepackage[colorlinks,linkcolor=blue,citecolor=blue]{hyperref}
\usepackage{xifthen}

\newcommand{\lP}{$\lambda$Prolog\xspace}
\newcommand{\ie}{i.e.,\xspace}
\newcommand{\eg}{e.g.,\xspace}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}
\newcommand{\blue}[1]{{\color[rgb]{0,0,1} #1}}

\newcommand{\DM}[1]{{\color[rgb]{0.23, 0.74, 0.23}\smallskip\par\noindent\textbf{DM: #1}}\smallskip}
\newcommand{\EP}[1]{\smallskip\par\noindent\red{#1}\smallskip}

\newcommand\proofsystem[1]{\mbox{\slshape #1}\xspace}
\newcommand\LK   {\proofsystem{LK}}
\newcommand\LJ   {\proofsystem{LJ}}
%% \newcommand\LKT  {\proofsystem{LKT}}
%% \newcommand\LKQ  {\proofsystem{LKQ}}
%% \newcommand\LKF  {\proofsystem{LKF}}
%% \newcommand\aLKF {\hbox{\proofsystem{LKF}\kern-2pt$^a$}\xspace}
%% \newcommand\LKU  {\proofsystem{LKU}}
%% \newcommand\LJT  {\proofsystem{LJT}}
%% \newcommand\LJQ  {\proofsystem{LJQ}}
%% \newcommand\LJF  {\proofsystem{LJF}}
%% \newcommand\aLJF {\hbox{\proofsystem{LJF}\kern-2pt$^a$}\xspace}
%% \newcommand\LKpos{\proofsystem{LKpos}}
%% \newcommand\LKneg{\proofsystem{LKneg}}
%% \newcommand\negclass[2]{\mathcal{N}_{#1}^{#2}}
%% \newcommand\posclass[2]{\mathcal{P}_{#1}^{#2}}

%% \newcommand{\tup}[1]{\langle #1\rangle}

\newcommand{\false}{f\/}
\newcommand{\imp}{\supset}
\newcommand{\nfalse}{f^-}
\newcommand{\ntrue}{t^-}
\newcommand{\pfalse}{f^+}
\newcommand{\ptrue}{t^+}
\newcommand{\ra}{\rightarrow}
\newcommand{\true }{t\/}
\newcommand{\veen}{\vee^{\!-}}
\newcommand{\veepn}{\vee^{\!\pm}}
\newcommand{\veep}{\vee^{\!+}}
\newcommand{\wedgen}{\wedge^{\!-}}
\newcommand{\wedgepn}{\wedge^{\!\pm}}
\newcommand{\wedgep}{\wedge^{\!+}}

\newcommand{\twoseq}[2]{#1\vdash #2}
\newcommand{\Twoseq}[3]{#1:#2\vdash #3}
\newcommand{\R}[2]{R #1 #2}

\begin{document}
\title{
%      A system of rules for sequent calculus\\(an abstract)\\
%      --- or ---\\
       Higher-level rules for sequent calculus\\(an abstract)
}
\author{Dale Miller\inst{1}%\orcidID{0000-0003-0274-4954}
        \and
        Elaine Pimentel\inst{2}%\orcidID{0000-0002-7113-0801}
}
\institute{
  Inria-Saclay \& LIX, Ecole Polytechnique, Palaiseau, France\\
%  \email{dale.miller@inria.fr}
  \and
  Dept. of Computer Science, University College London, UK \\
%  \email{e.pimentel@ucl.ac.uk}
}

\maketitle              

There has been a long-term interest in the trade-offs between
representing theories as collections of formulas accepted as axioms
and assumptions or as collections of inference rules that do not
involve logical connectives.

% Early work on an extreme perspective

In the setting of propositional intuitionistic logic, there was an early
paper on reducing all logical assumptions to inference rules.
Schroeder-Heister~\cite{schroeder-heister84} considered a
generalization of natural deduction where rules and not just formulas
can be assumed and discharged.  He showed that such rules are equally
expressive to Gentzen's natural deduction (where only formulas can be
assumed and discharged).  Later, Avron \cite{avron90ndjfl} reduced
this extended system to the sequent calculus presentation in which no
inference rules were assumed or discharged.  The LF
\cite{harper93jacm} specification framework generalized this work on
natural deduction to quantificational intuitionistic logic while
simultaneously encoding the extended natural deduction proofs using a
term representation within that dependently typed $\lambda$-calculus.

% Early work on replacing axioms with inference rules

Moving from natural deduction to sequent calculus permits treating
assumptions as either inference rules or as formulas in \emph{both}
intuitionistic and classical logic proofs.  In particular, it is
natural to ask whether or not an \LJ or \LK proof of the sequent
$\twoseq{\Gamma}{\Delta}$ can be built where the formulas in $\Gamma$
(the assumptions) are replaced with inference rules.  It was shown in
\cite{negri03aml} that when assumptions are \emph{geometric formulas},
this replacement is possible For example, if the multiset of
assumptions is of the form $\Gamma, \forall x\forall y\forall
z.[\R{x}{z}\imp\R{z}{y}\imp\R{x}{y}]$ then the assumption stating the
transitivity of the predicate $R$ can be replaced by one of the
following inference rules instead.
\[
  \infer[\hbox{backchain}]
        {\twoseq{\Gamma}{\R{x}{y}}}
        {\twoseq{\Gamma}{\R{x}{z}} & \twoseq{\Gamma}{\R{z}{y}}}
  \qquad
  \infer[\hbox{forwardchain}]
        {\twoseq{\Gamma,\R{x}{z},\R{z}{y}}{B}}
        {\twoseq{\Gamma,\R{x}{z},\R{z}{y},\R{x}{y}}{B}}
\]
In this way, we have replaced a formula with five occurrences of
logical connectives with one of two rules, neither of which contains
logical connectives.  If the only assumptions are, say, Horn clauses,
it is possible to have a complete proof system for atomic consequences
of such clauses using a proof system involving only atomic formulas.


While it is possible to generalize the restriction to geometric
formulas a bit to \emph{bipoles} by considering
polarity~\cite{marin22apal}, it is interesting to ask to what extent
can non-bipole formulas be replaced with rules in the sequent setting.
Such a question was addressed in \cite{negri16jlc}, where
\emph{generalized geometric formulas} are treated using a \emph{system
of rules}, a setting in which an inference rule can allow some of its
premises to have additional inference rules available.  These
additional inference rules are scoped over particular proofs of
premises.
%
We will show that such a scoping of inference rules is a direct
reading of inference rules in a polarized proof system.  For example,
the (polarized) formula
\[
  \forall x\forall y\exists z(x\le z\wedgep y\le z\wedgep\forall w(x\le w\wedgep y\le w\imp z\le w)) 
\]
yields the synthetic inference rule (terminology from
\cite{marin22apal})
\[
  \infer[(*)]
        {\Twoseq{\Sigma}{\Gamma}{\Delta}}
        {\Twoseq{\Sigma,z}{\forall w(x\le w\wedgep y\le w\imp z\le w),x\le z,y\le z,\Gamma}{\Delta}}
\]
Here, sequents are prefixed with a list of variables, \eg $\Sigma$,
that are the eigenvariables that may appear in the formulas of that
sequent: this prefix is intended to \emph{bind} variables over the
sequent it precedes.  Note that the prefix of the conclusion is
different from the prefix of the premise.  The availability of the
additional assumption $\forall w(x\le w\wedgep y\le w\imp z\le w)$ in
the premise corresponds exactly to having either one of the following
two inference rules scoped over that premise
\[
  \infer
        {\Twoseq{\Sigma,z}{\Gamma}{z\le w}}
        {\Twoseq{\Sigma,z}{\Gamma}{x\le w} & \twoseq{\Gamma}{y\le w}}
  \qquad
  \infer[,]
        {\Twoseq{\Sigma,z}{\Gamma,x\le w,y\le w}{\Delta}}
        {\Twoseq{\Sigma,z}{\Gamma,x\le w,y\le w, z\le w}{\Delta}}
\]
depending on whether or not the polarity of the $\le$ predicate is
negative or positive, respectively.  Thus, we can rewrite the
inference figure $(*)$ so that it does not mention any logical
connectives by stipulating that one of these two inference rules is
available to prove that premise.

It is also possible to directly relate 

The common implementation technique involving Tseitin constants are
often introduced in implementations of computational logic systems in
order to reduce the nesting of conjunctions and disjunctions \cite{???}
and implications \cite{???}.  We will point out how these constants are
used to reduce assumptions to collections of bipoles (at the cost of
introducing new predicate constants).

\bibliography{../../../Dropbox/Writing/references/master}
%\bibliography{extract} % Switch this with the line above

\bibliographystyle{splncs04}

% Formulas have their advantages: easier to formalize in proof
% assistants and they are more compact wrt non-determinism (dnf/cnf is
% needed generally to make inference rules).

\end{document}


% LocalWords:  Avron forwardchain Tseitin
